The Evolution of Phenotypically Invariant Genetic Networks

### 2017-08-31 kalman_walk_paper.tex
edited speciation section

  %where $A$ is a gene regulatory network (although this model can be generalized to other biological networks) -- a square matrix, 
  %and $B$ filters and translates the input to the system. The form of $B$ determines precisely how the state of the environment influences the internal gene network. $C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection. 
%
%
Furthermore, whereas the phenotype is a subset of molecules \emph{visible} to selection, the \emph{kryptotype} includes the molecular dynamics \emph{hidden} from selection. That is $\kappa(t)$ is a vector of the system's molecular concentrations at time $t$, directly, indrectly, and irrelevant to survival.
%
%The structure of $B$ and $C$ determines how systems process their environment and convert krypto- to phenotypes, respectively. 
%
\plr{This "Thus" needs expanding, e.g., "the solution to this equation is unique and given by" with a reference.}

@@

Gene regulatory networks with identical phenotypes do not necessarily have identical kryptotypes.
Any linear and minimal gene network 
-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes $(n=m)$
-- has an identical impulse response and thus phenotype given identical inputs $u(t)$, up to a change of coordinates. 
\plr{"has an identical phenotype" is not what you mean to say here}
\js{edited -- does that work?}
  \begin{equation}
    \begin{aligned}
      h(t) &= C e^{A t} B 
      = C V^{-1} V e^{A t} V^{-1} V B \\
      &= C V^{-1} e^{V A V^{-1} t} V B 
      = \bar{C} e^{\bar{A} t} \bar{B}
    \end{aligned}
  \end{equation}
Two gene networks, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, 
have the same phenotype, for all possible inputs, if they are related by a change of coordinates. 
\plr{is this the place to say "and only if, in the minimal dimension"?}

@@

%Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
 % \begin{align}
 %   A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
 % \end{align}
%and a \emph{Linear Evolutionary System} is, 
%\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
Therefore the set of all phenotypically identical (and minimal) systems can be parameterized as, $\mathcal{S}_{m}(V) := \{VA_0V^{-1}, VB_0, C_0V \}$,
  %  \begin{align} \label{eq:min_param}
  %    \mathcal{S}(V) := \left\{ \begin{array}{ll} V\dot{\kappa}(t) &= VA V^{-1} \kappa(t) + VB u(t) \\ \phi(t) &= C V^{-1} \kappa(t) \end{array} \right.
  %\end{align}
  where $V$ is any invertible matrix and it's elements are free parameters.

@@

More generally,
we denote by $\calA_n(\mathcal{S}_0)$ the set of all $n$-dimensional systems equivalent to $\mathcal{S}_0$:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \calA_n(\mathcal{S}_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  \\
      &= \left\{
        (A,B,C) : C A^r B = C_0 A_0^r B_0 \; \text{for}\; 1 \le r \le n-1 
      \right\} .
  \end{aligned}
\end{equation}
Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
\plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.

Two systems, even if in different dimensions, can have identical phenotypes if they are in $\calA_n$. Further, there is no unique triple $(A,B,C)$ for the mapping $u \mapsto \phi$. This implies that the gene regulatory network architecture per phenotype/environment pair is never unique -- that is \emph{all} gene regulatory networks can drift/rewire. This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition} \ref{apx:kalman}.

@@

>>>>

  %where $A$ is a gene regulatory network (although this model can be generalized to other biological networks) -- a square matrix, 
  %and $B$ filters and translates the input to the system. The form of $B$ determines precisely how the state of the environment influences the internal gene network. $C$ filters and translates the dynamics of the system and precesily determines the output, that is, what is visible to selection. 
%
%
Furthermore, whereas the phenotype is a subset of molecules \emph{visible} to selection, the \emph{kryptotype} includes the molecular dynamics \emph{hidden} from selection. That is $\kappa(t)$ is a vector of the system's molecular concentrations at time $t$.
%
%The structure of $B$ and $C$ determines how systems process their environment and convert krypto- to phenotypes, respectively. 
%
\plr{This "Thus" needs expanding, e.g., "the solution to this equation is unique and given by" with a reference.}

@@

Gene regulatory networks with identical phenotypes do not necessarily have identical kryptotypes.
Any linear and minimal gene network 
-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible genes $(n=\min)$
-- has an identical impulse response and thus phenotype given identical inputs $u(t)$, up to a change of coordinates. 
\plr{"has an identical phenotype" is not what you mean to say here}
\js{edited -- does that work?}
  \begin{equation}
    \begin{aligned}
      h(t) &= C e^{A t} B 
      = C V^{-1} V e^{A t} V^{-1} V B \\
      &= C V^{-1} e^{V A V^{-1} t} V B 
      = \bar{C} e^{\bar{A} t} \bar{B}
    \end{aligned}
  \end{equation}
Two gene networks, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, 
have the same phenotype, for all possible inputs, if they are related by a change of coordinates. 
\plr{is this the place to say "and only if, in the minimal dimension"?}

@@

%Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
 % \begin{align}
 %   A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
 % \end{align}
%and a \emph{Linear Evolutionary System} is, 
%\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
Therefore the set of all phenotypically identical (and minimal) systems can be parameterized as, $\mathcal{S}_{\min}(V) := \{VA_0V^{-1}, VB_0, C_0V \}$,
  %  \begin{align} \label{eq:min_param}
  %    \mathcal{S}(V) := \left\{ \begin{array}{ll} V\dot{\kappa}(t) &= VA V^{-1} \kappa(t) + VB u(t) \\ \phi(t) &= C V^{-1} \kappa(t) \end{array} \right.
  %\end{align}
  where $V$ is any invertible matrix and it's elements are free parameters.

@@

More generally,
we denote by $\calA_n(\mathcal{S}_0)$ the set of all $n$-dimensional systems equivalent to $\mathcal{S}_0$:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
    \calA_n(\mathcal{S}_0) 
      &= \left\{
        (A,B,C) : C e^{At} B = C_0 e^{A_0 t} B_0 \; \text{for}\; t \ge 0 
      \right\}  \\
      &= \left\{
        (A,B,C) : C A^r B = C_0 A_0^r B_0 \; \text{for}\; 1 \le r \le n-1 
      \right\} .
  \end{aligned}
\end{equation}
  \begin{align}
    \mathcal{K} := \bigcup_{n=\min}^{\infty} \calA_n(\calA_0)
  \end{align}
Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
\plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.

Two systems, even if in different dimensions, will have identical phenotypes if they are in $\mathcal{K}$. Further, there is no unique triple $(A,B,C)$ for the mapping $u \mapsto \phi$. This implies that the gene regulatory network architecture per phenotype/environment pair is never unique -- that is \emph{all} gene regulatory networks can drift/rewire. This set can be precisley defined and completely parameterized using the \emph{Kalman decomposition} \ref{apx:kalman}.

A large, adapted population will move around in $\mathcal{K}$ given environmental and selective stasis. Even if there is a conservation of phenotype through evolutionary time, the molecular mechanism underpinning the phenotype is not constrained and likely will not be conserved.

@@

\paragraph{Kryptotypic variation and hybrid incompatibility}
Phenotypically identical yet kryptotypically variant gene networks, when brought together and/or recombined during sexual reproduction, can lead to phenotypic inviability.

For instance, the phenotypic dynamics of a diploid organism are a function of its two haploid genomes $x$ and $y$: $P(x,y) = \left(x + y\right)/2$. Additionally, during meiosis, a diploid recombines its two genomes to form a new one: $\left(x \times y \right) \rightarrow z$; by randomly swapping matrix rows between two systems $\mathcal{S}$ and $\mathcal{S}'$.

Even if two genomes $x$ and $y$ are both in $\mathcal{K}$, their diploid systems $P(x,y)$, and their recombinant systems $P(z,z)$, may not be in $\mathcal{K}$. This is because $\mathcal{K}$ is not mathematically required to be closed under averaging and recombination.

If sexual recombination among systems drawn from $\mathcal{K}$ yields systems with divergent phenotypes, populations containing significant diversity in $\mathcal{K}$ will carry a large genetic load, and isolated populations may fail to produce hybrids with viable phenotypes.
